Synthesising Concise Termination and Non-Termination Proofs for Bit-Vector Programs

### 2014-07-08 papers/termination/paper.tex
fix [CT] def
Soundness proofs

@@

% All we need to do in order to use {\bf [CT]} in any context %environment and regardless of the complexity of the loops's body 
% is generate constraints over-approximating %abstracting 
% the environment and the loop's body. %: for termination, we need over-approximating abstractions, whereas for non-termination 
As the generation of such constraints is standard and covered by several other works \cite{DBLP:conf/pldi/GrebenshchikovLPR12,DBLP:conf/pldi/GulwaniSV08}, 
we will not provide the full algorithm, but rather illustrate it through the example in Figure~\ref{fig:environment-model}.
Our technique proves the termination of each loop separately, 
with the exception of nested loops, whose termination we encode in the same formula. For the current example, the termination formula is given on the right side of 
Figure~\ref{fig:environment-model}:
%% Following the same design, let us assume that we have already proven the termination of loop $L_1$, and we are now trying to prove that $L_2$ and $L_3$ terminate, which we encode
%% as the conjunction of constraints on the right side of the figure:
\begin{itemize}
\item $T_o$  over-approximates $L_1$'s transition relation.
\item $P$ is a summary of $L_2$ that over-approximates the transitive closure of its transition relation.
\item If the formula is satisfiable, then $R_1$ and $R_2$ are ranking functions for $L_1$ and $L_2$, respectively.
\end{itemize}

%% Note that despite its intricacy, the constraint system is still a formula in the synthesis fragment.
%% Since the outer loop of this program does indeed terminate, the system is satisfiable, as witnessed by:
%% \begin{align*}
%% W_1(\langle v, x, y, z \rangle ) & \triangleq x \leq y + 1 \\
%% W_2(\langle v, x, y, z \rangle ) & \triangleq x = y + 1 \\
%% S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) & \triangleq x ' = x \wedge y' = y \wedge \\
%% & z' + v' = x' \wedge v' \leq y' \\
%% R_2(x) & = x\\
%% R_3 & = ..
%% \end{align*}

\begin{figure*}
\begin{framed}
\begin{minipage}{0.16\textwidth}
\begin{lstlisting}[mathescape=true]
$L_1:$
while (i<n){
  j = 0;

$L_2:$
  while (j$\leq$i){
    j = j + 1;
  }

  i = i + 1;
}
\end{lstlisting}
\end{minipage}
\vline
\begin{minipage}{0.84\textwidth}
\begin{align*}
 \exists T_o, P, R_1, R_2 . \forall i, j, n, i', j', n' . \\
   %% x = 0 & \rightarrow W_1(v, x, y, z) \, \wedge \\
   %% x \leq y \wedge W_1(v, x, y, z) & \rightarrow W_1(v, x+1, x, y, z) \, \wedge \\
   %% x > 0 \wedge x > y \wedge W_1(v, x, y, z) & \rightarrow W_2(v, x, y, z) \, \wedge \\
   P(\langle i, j, n \rangle, \langle i', j', n' \rangle) \wedge j' \leq i' & \rightarrow P(\langle i, j, n \rangle, \langle i', j'+1, n' \rangle) \, \wedge \\
   & R_2(i,j,n) > 0 \, \wedge \\
   & R_2(i,j,n) > R_2(i',j'+1,n') \, \wedge\\
   i < n \wedge T_o(i, j, n) & \rightarrow P(\langle i, j, n \rangle, \langle i, 0, n \rangle) \, \wedge \\
   i < n \wedge T_o(i, j, n) \wedge P(\langle i, j, n \rangle, \langle i', j', n' \rangle) \wedge j' > i' & \rightarrow \\
   & T_o(i'+1, j', n') \, \wedge \\
   & R_1(i, j, n) > 0 \, \wedge \\
   & R_1(i, j, n) > R_1(i'+1, j', n')
\end{align*}
\end{minipage}
\end{framed}
\caption{A program with nested loops and its termination formula\label{fig:environment-model}}
\end{figure*}

%% \begin{figure*}
%% \begin{framed}
%% \begin{minipage}{0.145\textwidth}
%% \begin{lstlisting}[mathescape=true]
%% x = 0;
%% $L_1:$
%% while(x<=y)
%%   x++;

%% $L_2:$
%% while(x>0){
%%   z = x;
%%   v = 0;

%% $L_3:$
%%   while(v<y){
%%     z--;
%%     v++;
%%   }

%%   x -= z;
%%   y -= z;
%% }
%% \end{lstlisting}
%% \end{minipage}
%% \vline
%% \begin{minipage}{0.85\textwidth}
%% \begin{align*}
%%  \exists W_1, W_2, R_2, R_3, S . \forall v, x, y, z, v', x', y', z' . \\
%%    x = 0 & \rightarrow W_1(v, x, y, z) \, \wedge \\
%%    x \leq y \wedge W_1(v, x, y, z) & \rightarrow W_1(v, x+1, x, y, z) \, \wedge \\
%%    x > 0 \wedge x > y \wedge W_1(v, x, y, z) & \rightarrow W_2(v, x, y, z) \, \wedge \\
%%    S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) \wedge v' < y' & \rightarrow S(\langle v, x, y, z \rangle, \langle v'+1, x', y', z'-1 \rangle) \, \wedge \\
%%    & R_3(v,x,y,z) > 0 \, \wedge \\
%%    & R_3(v,x,y,z) > R_3(v'+1,x',y',z'-1) \, \wedge\\
%%    x > 0 \wedge W_2(v, x, y, z) & \rightarrow S(\langle v, x, y, z \rangle, \langle 0, x, y, x \rangle) \, \wedge \\
%%    x > 0 \wedge W_2(v, x, y, z) \wedge S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) \wedge v' \geq y' & \rightarrow \\
%%    & W_2(v', x' - z', y' - z', z') \, \wedge \\
%%    & R_2(v, x, y, z) > 0 \, \wedge \\
%%    & R_2(v, x, y, z) > R_2(v', x' - z', y'-z', z')
%% \end{align*}
%% \end{minipage}
%% \end{framed}
%% \caption{A non-trivial program and its termination formula\label{fig:environment-model}}
%% \end{figure*}

@@

\subsection{Generalised Termination and Non-termination Formula}
At this point, we know how to construct two formulae for a loop $L$: one that is
satisfiable iff $L$ is terminating and another that is
satisfiable iff it is non-terminating.  We will call these formulae $\phi$ and $\psi$
respectively:
\begin{align*}
 \exists P_N . \forall x, x' . \phi(W_N, x, x') \\
 \exists P_T . \forall x, x' . \psi(W_T, x, x')
\end{align*}
We can combine these:
\begin{align*}
 (\exists P_N . \forall x, x'. \phi(W_N, x, x')) \vee (\exists P_T . \forall x, x', \psi(W_T, x, x'))
\end{align*}
Which simplifies to:
\begin{definition}[Generalised Termination Formula]
\begin{align*}
 \exists P_N, P_T. \forall x, x', y, y', \phi(P_N, x, x') \vee \psi(P_T, y, y') 
\end{align*}
\end{definition}

Since $L$ either terminates or does not terminate, this formula is a tautology in the synthesis fragment.
A solution to the formula would include witnesses $P_N$ and $P_T$, which are putative proofs of non-termination
and termination respectively.  Exactly one of these will be a genuine proof, so we can check
first one and then the other.

@@

\begin{algorithm*}
 \caption{Abstract refinement algorithm
 \label{alg:cegis}}
 \vspace{-1.5em}
 \begin{multicols}{2}
 \begin{algorithmic}[1]
 \Statex
\Function{synth}{{\sc inputs}}
  \Let{$(i_1, \ldots, i_N)$}{{\sc inputs}}
  \Let{query}{$\exists P, x_0 . \forall i \in \text{\sc inputs} . \sigma(P, x_0, i)$}
  \Let{result}{decide(query)}
  \If{result.satisfiable}
    \State \Return{result.model}
  \Else
    \State \Return{unsatisfiable}
  \EndIf
\EndFunction
\Statex
\Function{verif}{P}
  \Let{query}{$\exists x . \lnot \sigma(P, x_0, x)$}
  \Let{result}{decide(query)}
  \If{result.satisfiable}
    \State \Return{result.model}
  \Else
    \State \Return{valid}
  \EndIf
\EndFunction
\columnbreak
\Statex
\Function{RefinementLoop}{}
  \Let{{\sc inputs}}{$\emptyset$}
  \Loop
    \Let{candidate}{\Call{synth}{{\sc inputs}}}
    \If{candidate = UNSAT}
      \State \Return{unsatisfiable}
    \EndIf
    \Let{res}{\Call{verif}{candidate}}
    \If{res = valid}
      \State \Return{candidate}
    \Else
      \Let{{\sc inputs}}{{\sc inputs} $\cup$ res}
    \EndIf
  \EndLoop
\EndFunction
 \end{algorithmic}
 \end{multicols}
\end{algorithm*}

@@

Termination for bit-vector programs is known to be
PSPACE-complete~\cite{DBLP:conf/tacas/CookKRW10}.  We first record the
expected results for soundness and completeness of our algorithm for
deciding it.
%
\begin{theorem}\label{thm:synth-sound}
Algorithm~\ref{alg:cegis} is sound -- if it terminates with witness $(P, x_0)$, then
$(P, x_0) \models \sigma$.
\end{theorem}

\begin{proof}
 The procedure {\sc RefinementLoop} terminates only if {\sc Verif} returns ``valid''.  In that
 case, $\exists x . \lnot \sigma(P, x_0, x)$ is unsatisfiable and so $\forall x . \sigma(P, x_0, x)$ holds.
\end{proof}

\begin{theorem}
 \label{thm:synth-semicomplete}
 If the existential first-order theory used to express the specification $\sigma$ is decidable,
 Algorithm~\ref{alg:cegis} is semi-complete -- if a solution $(P, x_0) \models \sigma$
 exists then Algorithm~\ref{alg:cegis} will terminate.  However, if no program
 satisfies the specification, the algorithm may not terminate.  Furthermore if the domain of inputs $X$
 is finite, the algorithm is guaranteed to terminate.
\end{theorem}

\begin{proof}
 Since the {\sc ExplicitSearch} routine enumerates all programs (as can be seen by induction on
 the program length $l$), it will eventually enumerate a program that meets the specification
 in whatever set of inputs are currently being tracked.  Since the first-order theory is
 decidable, the query in {\sc Verif} will succeed for this program, causing the algorithm to terminate.
 The set of correct programs is therefore recursively enumerable and Algorithm~\ref{alg:cegis}
 enumerates this set, so it is semi-complete.

 If the domain $X$ is finite then the loop in procedure {\sc RefinementLoop} can only
 iterate $\| X \|$ times, since by this time all of the elements of $X$ would have been
 added to the inputs set.  Therefore if the {\sc Synth} procedure always terminates (which it does
 by assumption of decidability), the whole algorithm terminates.
\end{proof}

@@

@@

\begin{proof}
A very inefficient construction which computes $f$ is:
%
\begin{verbatim}
t1 = f(0)
t2 = v1 == 1
t3 = ITE(t2, f(1), t1)
t4 = v1 == 2
t5 = ITE(t4, f(2), t3)
...
\end{verbatim}
%
Thus, every $f$ is computed a program of size at most $b \times
2^{bk + 1}$.

(Tidy this up).
\end{proof}

@@

\begin{proof}
 A non-termination proof is a triple $\langle N, C, x_0 \rangle$ where $N \subseteq \mathcal{S}$ is
 a (finite) recurrence set and $C : \mathcal{S} \to \mathcal{S}$ is a Skolem function choosing
 a successor for each $x \in N$.  Since $\mathcal{S}$ is finite, and by Lemma~\ref{lem:l-func}, both
 $N$ and $C$ are computed by finite $\mathcal{L}$-programs.
\end{proof}

@@

\begin{proof}
 By Theorem~\ref{thm:l-term}, if $L$ terminates then there exists a termination proof $P_T$ expressible
 in $\mathcal{L}$.  Since $\phi$ is a termination specification, $P_T \models \phi$ and so $P_T \models \phi \vee \psi$.

 Similarly if $L$ does not terminate for some input, by Theorem~\ref{thm:l-nonterm} there is a non-termination
 proof $P_N$ expressible in $\mathcal{L}$, so $P_N \models \psi$ and so $P_N \models \phi \vee \psi$. 
\end{proof}

\begin{theorem}
 Algorithm~\ref{alg:cegis} terminates on the generalised termination specification of Definition~\ref{def:general-termination}.
\end{theorem}

\begin{proof}
 The specification is satisfiable (Thoerem~\ref{thm:generalised-sat}) and Algorithm~\ref{alg:cegis} is semi-complete
 (Theorem~\ref{thm:synth-semicomplete}).
\end{proof}

@@

{\bf TODO:} Tie everything up into one algorithm (build the formula, invoke CEGIS, check which witness is
correct) then proof soundness for that.

>>>>

\begin{theorem}
 \label{thm:snt}
 Formula {\bf [ONT]} (Definition~\ref{def:ont}) is satisfiable iff {\bf [SNT]} (Definition~\ref{def:snt}) is.
\end{theorem}

@@

% All we need to do in order to use {\bf [CT]} in any context %environment and regardless of the complexity of the loops's body 
% is generate constraints over-approximating %abstracting 
% the environment and the loop's body. %: for termination, we need over-approximating abstractions, whereas for non-termination 
As the generation of such constraints is standard and covered by several other works \cite{DBLP:conf/pldi/GrebenshchikovLPR12,DBLP:conf/pldi/GulwaniSV08}, 
we will not provide the full algorithm, but rather illustrate it through the example in Figure~\ref{fig:environment-model}.
Our technique proves the termination of each loop separately and in program order, 
with the exception of nested loops, whose termination we encode in the same formula.
Following the same design, let us assume that we have already proven the termination of loop $L_1$, and we are now trying to prove that $L_2$ and $L_3$ terminate, which we encode
as the conjunction of constraints on the right side of the figure:
%% =======
%% %
%% As the generation of such constraints is standard and covered by several
%% other
%% works~\cite{DBLP:conf/pldi/GrebenshchikovLPR12,DBLP:conf/pldi/GulwaniSV08},
%% we will not provide the full algorithm, but rather illustrate it through the
%% example in Figure~\ref{fig:environment-model}.  Let us assume that we have
%% already proven the termination of loops $L_1$ and $L_3$, and we are now
%% trying to prove that $L_2$ terminates, which we encode as the conjunction of
%% constraints on the right-hand side of the figure:
%% %
%% >>>>>>> ee2c21bbc574ab36e7d6739c0e82c7a03d930789
\begin{itemize}
\item $W_1$ and $W_2$ are (over-approximating) inductive invariants for $L_1$ and $L_2$, respectively.
\item $S$ is a summary of $L_3$ that over-approximates the transitive closure of its transition relation.
\item If the formula is satisfiable, then $R_2$ and $R_3$ are ranking functions for $L_2$ and $L_3$, respectively.
\end{itemize}

Note that despite its intricacy, the constraint system is still a formula in the synthesis fragment.
Since the outer loop of this program does indeed terminate, the system is satisfiable, as witnessed by:
\begin{align*}
W_1(\langle v, x, y, z \rangle ) & \triangleq x \leq y + 1 \\
W_2(\langle v, x, y, z \rangle ) & \triangleq x = y + 1 \\
S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) & \triangleq x ' = x \wedge y' = y \wedge \\
& z' + v' = x' \wedge v' \leq y' \\
R_2(x) & = x\\
R_3 & = ..
\end{align*}

\begin{figure*}
\begin{framed}
\begin{minipage}{0.145\textwidth}
\begin{lstlisting}[mathescape=true]
x = 0;
$L_1:$
while(x<=y)
  x++;

$L_2:$
while(x>0){
  z = x;
  v = 0;

$L_3:$
  while(v<y){
    z--;
    v++;
  }

  x -= z;
  y -= z;
}
\end{lstlisting}
\end{minipage}
\vline
\begin{minipage}{0.85\textwidth}
\begin{align*}
 \exists W_1, W_2, R_2, R_3, S . \forall v, x, y, z, v', x', y', z' . \\
   x = 0 & \rightarrow W_1(v, x, y, z) \, \wedge \\
   x \leq y \wedge W_1(v, x, y, z) & \rightarrow W_1(v, x+1, x, y, z) \, \wedge \\
   x > 0 \wedge x > y \wedge W_1(v, x, y, z) & \rightarrow W_2(v, x, y, z) \, \wedge \\
   S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) \wedge v' < y' & \rightarrow S(\langle v, x, y, z \rangle, \langle v'+1, x', y', z'-1 \rangle) \, \wedge \\
   & R_3(v,x,y,z) > 0 \, \wedge \\
   & R_3(v,x,y,z) > R_3(v'+1,x',y',z'-1) \, \wedge\\
   x > 0 \wedge W_2(v, x, y, z) & \rightarrow S(\langle v, x, y, z \rangle, \langle 0, x, y, x \rangle) \, \wedge \\
   x > 0 \wedge W_2(v, x, y, z) \wedge S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) \wedge v' \geq y' & \rightarrow \\
   & W_2(v', x' - z', y' - z', z') \, \wedge \\
   & R_2(v, x, y, z) > 0 \, \wedge \\
   & R_2(v, x, y, z) > R_2(v', x' - z', y'-z', z')
\end{align*}
\end{minipage}
\end{framed}
\caption{A non-trivial program and its termination formula\label{fig:environment-model}}
\end{figure*}

@@

\subsection{Generalised Termination and Non-termination Formula}
At this point, we know how to construct two formulae for a loop $L$: one that is
satisfiable iff $L$ is terminating and another that is
satisfiable iff it is non-terminating.  We will call these formulae $\phi$ and $\psi$
respectively:
\begin{align*}
 \exists P_T . \forall x, x' . \phi(P_T, x, x') \\
 \exists P_N . \forall x . \psi(P_N, x)
\end{align*}
We can combine these:
\begin{align*}
 (\exists P_T . \forall x, x'. \phi(P_T, x, x')) \vee (\exists P_N . \forall x, \psi(P_N, x))
\end{align*}
Which simplifies to:
\begin{definition}[Generalised Termination Formula {\bf [GT]}]
\label{def:general-termination}
\begin{align*}
 \exists P_T, P_N. \forall x, x', y, \phi(P_T, x, x') \vee \psi(P_N, y)
\end{align*}
\end{definition}

Since $L$ either terminates or does not terminate, this formula is a tautology in the synthesis fragment.
A solution to the formula would include witnesses $P_N$ and $P_T$, which are putative proofs of non-termination
and termination respectively.  Exactly one of these will be a genuine proof, so we can check
first one and then the other, as shown in Algorithm~\ref{alg:termination}.

\begin{algorithm}
 \caption{Bit-vector termination
 \label{alg:termination}}
 %\vspace{-1.5em}
 \begin{algorithmic}[1]
 \Statex
\Function{terminates}{$L$}
  %\Let{termination-spec}{$\exists W, R. \forall x, x' . \phi(W, R, x, x')$} \Comment{Formula {\bf [CT]}}
  %\Let{nonterm-spec}{$\exists N, C, x_0, \forall x. \psi(N, C, x_0, x)$} \Comment{Formula {\bf [SNT]}}
  \Statex  \Comment{Formula {\bf[GT]}:}
  \Let{spec}{$\exists P_T, P_N. \phi(P_T, x, x') \vee \psi(P_N, y)$}
  \Let{$(P_T, P_N)$}{{\sc Synthesise}(spec)}
  \If{decide($\exists x, x' . \lnot \phi(P_T, x, x'))$ = unsat}
    \State \Return{$L$ terminates}
  \ElsIf{decide($\exists x . \lnot \psi(P_N, x))$ = unsat}
    \State \Return{$L$ does not terminate}
  \EndIf
\EndFunction
 \end{algorithmic}
\end{algorithm}

@@

\begin{algorithm*}
 \caption{Abstract refinement algorithm
 \label{alg:cegis}}
 \vspace{-1.5em}
 \begin{multicols}{2}
 \begin{algorithmic}[1]
 \Statex
\Function{synth}{{\sc inputs}, $\sigma$}
  \Let{$(i_1, \ldots, i_N)$}{{\sc inputs}}
  \Let{query}{$\exists P, x_0 . \forall i \in \text{\sc inputs} . \sigma(P, x_0, i)$}
  \Let{result}{decide(query)}
  \If{result.satisfiable}
    \State \Return{result.model}
  \Else
    \State \Return{unsatisfiable}
  \EndIf
\EndFunction
\Statex
\Function{verif}{$P, x_0, \sigma$}
  \Let{query}{$\exists x . \lnot \sigma(P, x_0, x)$}
  \Let{result}{decide(query)}
  \If{result.satisfiable}
    \State \Return{result.model}
  \Else
    \State \Return{valid}
  \EndIf
\EndFunction
\columnbreak
\Statex
\Function{Synthesise}{$\sigma$}
  \Let{{\sc inputs}}{$\emptyset$}
  \Loop
    \Let{candidate}{\Call{synth}{{\sc inputs}}}
    \If{candidate = UNSAT}
      \State \Return{unsatisfiable}
    \EndIf
    \Let{res}{\Call{verif}{candidate}}
    \If{res = valid}
      \State \Return{candidate}
    \Else
      \Let{{\sc inputs}}{{\sc inputs} $\cup$ res}
    \EndIf
  \EndLoop
\EndFunction
 \end{algorithmic}
 \end{multicols}
\end{algorithm*}

@@

Termination for bit-vector programs is known to be
PSPACE-complete~\cite{DBLP:conf/tacas/CookKRW10}.  We first record the
expected results for soundness and completeness of our algorithm for
deciding it.  Our strategy for proving soundness and completeness
is first to prove that Algorithm~\ref{alg:cegis} (as instanciated
with \newC as a background logic) is a semi-decision
procedure.  Next, we will show that $\mathcal{L}$ is expressive enough
to compute (non-)termination proofs for every bit-vector program.
We will then combine these results to show that the procedure as a whole
terminates with a valid proof for every input program.
%
\begin{theorem}\label{thm:synth-sound}
Algorithm~\ref{alg:cegis} is sound -- if it terminates with witness $(P, x_0)$, then
$(P, x_0) \models \sigma$.
\end{theorem}

\begin{proof}
 The procedure {\sc Synthesise} terminates only if {\sc Verif} returns ``valid''.  In that
 case, $\exists x . \lnot \sigma(P, x_0, x)$ is unsatisfiable and so $\forall x . \sigma(P, x_0, x)$ holds.
\end{proof}

\begin{theorem}
 \label{thm:synth-semicomplete}
 If the existential first-order theory used to express the specification $\sigma$ is decidable
 and the domain of inputs $X$ is finite,
 Algorithm~\ref{alg:cegis} is semi-complete -- if a solution $(P, x_0) \models \sigma$
 exists then Algorithm~\ref{alg:cegis} will terminate.  However, if no program
 satisfies the specification, the algorithm may not terminate.
\end{theorem}

\begin{proof}
 If the domain $X$ is finite then the loop in procedure {\sc Synthesise} can only
 iterate $\| X \|$ times, since by this time all of the elements of $X$ would have been
 added to the inputs set.  Therefore if the {\sc Synth} procedure always terminates,
 Algorithm~\ref{alg:cegis}.

 Since the {\sc ExplicitSearch} routine enumerates all programs (as can be seen by induction on
 the program length $l$), it will eventually enumerate a program that meets the specification
 in whatever set of inputs are currently being tracked, since by assumption such a program
 exists.  Since the first-order theory is
 decidable, the query in {\sc Verif} will succeed for this program, causing the algorithm to terminate.
 The set of correct programs is therefore recursively enumerable and Algorithm~\ref{alg:cegis}
 enumerates this set, so it is semi-complete.
\end{proof}

@@

We will now show that $\mathcal{L}$ is expressive enough to encode (non-)termination proofs for
every bit-vector program.

@@

\begin{proof}
Without loss of generality, let $X = Y = \mathbb{N}_b^k$ the set of
$k$-tuples of natural numbers less than $b$.
A very inefficient construction which computes the first co-ordinate
of the output $y$ is:
%
\begin{verbatim}
t1 = f(0)
t2 = v1 == 1
t3 = ITE(t2, f(1), t1)
t4 = v1 == 2
t5 = ITE(t4, f(2), t3)
...
\end{verbatim}
%
Where the \verb|f(n)| are literal constants that are to appear in the program text.
This program is of length $2b - 1$, and so all $k$ co-ordinates of the output $y$
are computed by a program of size at most $2bk - k$.
\end{proof}

@@

\begin{proof}
 A non-termination proof is a triple $\langle N, C, x_0 \rangle$ where $N \subseteq \mathcal{S}$ is
 a (finite) recurrence set and $C : \mathcal{S} \to \mathcal{S}$ is a Skolem function choosing
 a successor for each $x \in N$.  Since $\mathcal{S}$ is finite, so by Lemma~\ref{lem:l-func} both
 $N$ and $C$ are computed by finite $\mathcal{L}$-programs.
\end{proof}

@@

\begin{proof}
 For any $P, P', \sigma, \sigma$, if $P \models \sigma$ then $(P, P') \models \sigma \vee \sigma'$.

 By Theorem~\ref{thm:l-term}, if $L$ terminates then there exists a termination proof $P_T$ expressible
 in $\mathcal{L}$.  Since $\phi$ is a termination specification, $P_T \models \phi$ and so
 for any $P_N$, $(P_T, P_N) \models \phi \vee \psi$.

 Similarly if $L$ does not terminate for some input, by Theorem~\ref{thm:l-nonterm} there is a non-termination
 proof $P_N$ expressible in $\mathcal{L}$, so $P_N \models \psi$ and hence for any $P_T$, $(P_T, P_N) \models \phi \vee \psi$.

 So in either case ($L$ terminates or does not), there is a witness in $\mathcal{L}$ satisfying $\phi \vee \psi$.
\end{proof}

\begin{theorem}
 Algorithm~\ref{alg:termination} is sound and complete -- it terminates for all input loops $L$ with
 a correct termination verdict.
\end{theorem}

\begin{proof}
 The background logic for Algorithm~\ref{alg:termination} is \newC, for which safety is decidable.  Also, the set of
 inputs is just the state space of $L$, which is finite.  Therefore Theorems~\ref{thm:generalised-sat} and \ref{thm:synth-semicomplete}
 hold -- the specification \lstinline!spec! is satisfiable and the instanciation of Algorithm~\ref{alg:cegis} is semi-complete.
 Therefore the call to {\sc Synthesise} terminates with a witness $(P_T, P_N)$.
 If $L$ terminates then $P_T$ is a termination proof (Theorem~\ref{thm:ct}).  Otherwise $P_N$ is a non-termination proof
 (Theorem~\ref{thm:snt}).  Exactly one of these cases holds, and the background theory being decidable means that each
 call to \lstinline|decide()| will terminate, which allows us to determine which of $P_T$ and $P_N$ is a correct proof.
\end{proof}

@@
